Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-03-22 PAPER.tex
made syntax/typo changes from full-read-through up to Section 3

Bootstrapping uses sampling with replacement to estimate a posterior distribution. For example, measurements (like power 
spectra) can be made from different samples of data. Each of these measurements is a different realization drawn from some underlying distribution, and realizations are correlated with each other to a degree set by the fraction of sampled points that are held in common 
between them. Through the process of resampling and averaging along different axes, we can estimate error bars for 
our results which represent the underlying distribution of values that are allowed by our measurements (\citealt{efron_tibshirani1994}; \citealt{andrae2010}).

Suppose we have $N$ different measurements targeting the same quantity ($N$ power spectrum measurements made along different axes, such as baselines or times). 
Bootstrapping means that we form $N_{boot}$ (often a large number) bootstraps, where each bootstrap is a random selection 
of the $N$ measurements. Bootstraps each have dimensions of $N$, and the values populated into each bootstrap are drawn 
from the original set of measurements with replacement (i.e. every $n^{th}$ slot in $N$ is filled randomly for each bootstrap). Next we take 
the mean of each bootstrap to collapse it from an array of length $N$ to a single number (we are interested in the mean statistic 
here, but any function of interest can be applied to each bootstrap as long as it's the same function for each one). The error (on 
the mean) is then computed as the standard deviation across all bootstraps.

@@

This example highlights the importance of understanding how analysis techniques (e.g. fringe-rate filtering) can affect a 
common statistical procedure like bootstrapping. Bootstrapping as a means of estimating power spectrum errors from real 
fringe-rate filtered data requires knowledge of the number of independent samples, which is not always a trivial task. For 
example, computing the effective number of independent samples of fringe-rate filtered data is not as simple as counting the 
number of averages performed. Down-sampling a time-averaged signal is straightforward using a boxcar average, but non-trivial with a more complicated convolution function that has long tails. Hence, we do not recommend bootstrapping unless the 
number of independent samples along the axis that is being re-sampled is well-determined. In Section \ref{sec:Boot}, we explain how our bootstrapping procedure has changed in PAPER's analysis in light of this issue.

@@

\begin{comment} % REMOVED THIS HALF OF BOOTSTRAPPING SECTION
We will now discuss a second subtle feature of bootstrapping that fails to maximize the sensitivity of a dataset. Suppose we 
have $5$ independent measurements of the sky (from $5$ different baselines, for example). A bootstrap then consists of $5$ 
measurements that are drawn randomly with replacement from the original set. If all $5$ spaces are filled randomly, there is a 
high probability that some measurements will be repeated in the bootstrap because they are drawn more than once. The 
bootstrap may therefore consist of only $3$ or $4$ independent measurements --- a number smaller than the total number of 
samples. In fact, the probability of drawing $5$ completely independent measurements is less than $4\%$.

In order to maximize sensitivity, we desire as many independent samples as possible. However, solely using all $5$ 
independent measurements does not allow variation between bootstraps.

Therefore, we use a slightly modified bootstrapping method. For each bootstrap, we draw $N-1=4$ samples without 
replacement and draw the last slot randomly with replacement. In doing so there is a small chance a bootstrap will consist of 
$5$ independent measurements, but even with $4$ our sensitivity is nearly maximized. 
%One may wonder whether this change is still a legitimate way of error estimating since random sampling only occurs for one value in a bootstrap. However, as long as the number of possible variations is greater than the number of bootstraps we perform, it is a valid way to uncover the inherent variability in a dataset.

In Figure \ref{fig:toy_error2} we compare these two methods of bootstrapping: sampling all elements randomly (black points) 
versus sampling just the final element randomly (gray points). The two converge for datasets with small numbers of elements, 
when filling a few spots randomly is nearly the same as filling only the last spot randomly. However, there is also increased 
scatter in this regime due to small number statistics.

The difference between the two methods is most pronounced for datasets with large numbers of elements. As a dataset grows 
in size, it becomes increasingly rare to draw entirely independent samples, and thus the benefit of ensuring, say, $999$ 
independent samples out of $1000$ is most noticeable. The more elements in our dataset, the more sensitivity we can gain by 
mandating that most of our measurements are independent ones. An analog for the y-axis in Figure \ref{fig:toy_error2} is 
therefore the sensitivity of a final power spectrum measurement.
\end{comment}

@@

In Section \ref{sec:SiglossOverview}, we discussed signal loss as a form of multiplicative bias to estimates of the signal. Foregrounds are another type of bias, but an additive one instead of multiplicative. Foreground bias is perhaps one of the main factors limiting 21\,cm results, as foreground signals lie $\sim4$-$5$ orders of 
magnitude above the cosmological signal. Though there are many techniques proposed for removing foregrounds (see e.g. \citealt{vedantham_et_al2012}; \citealt{parsons_et_al2012a}; \citealt{parsons_et_al2012b}; \citealt{dillon_et_al2013a}; \citealt{wang_et_al2013}; \citealt{parsons_et_al2014}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}; \citealt{dillon_et_al2015}; \citealt{pober_et_al2016a}; \citealt{trott_et_al2016}), most 
experiments currently remain limited by residuals rather than noise, especially at low $k$.

@@

Although smooth spectrum foregrounds preferentially show up at low delay, or low $k$ modes, their isolation within the wedge is not perfect. In deep measurements, power spectrum measurements at $k_{\parallel}$ values beyond 
the delay associated with the length of a baseline are often still contaminated at a low level. This leakage, particularly at low $k$'s, can be attributed to 
convolution kernels associated with Fourier-transforming visibilities into delay-space. In other words, smooth-spectrum 
foregrounds appear as $\delta$-functions in delay-space, convolved by the Fourier transform of the source spectrum, the signal chain, and the 
antenna response, all of which could smear out the foregrounds and cause leakage outside the wedge \citep[e.g.][]{ewall-wice_et_al2017}.

@@

Determining the source of positive non-EoR detections at higher $k$'s is more difficult. In previous power spectrum results, these detections have been explained as instrumental systematics, particularly time-variable cross talk, RFI, cable reflections, and calibration errors (\citetalias{ali_et_al2015}; \citealt{parsons_et_al2014}; \citealt{dillon_et_al2014}; \citealt{beardsley_et_al2016}; \citealt{patil_et_al2017}). In the next section (Section \ref{sec:JackknifeOverview}), we will present some tests that can help distinguish these excesses from that of EoR.

In addition to foreground bias, noise can also be responsible for positive power spectrum detections if thermal noise is 
multiplied by itself. Every 21\,cm visibility measurement contains thermal noise that is comprised of receiver and sky noise. 
We expect this noise to be independent between antennas and thus we can beat it down (increase sensitivity) by integrating 
longer, using more baselines, etc. However, the squaring of noise occurs when cross-multiplying visibilities, which is shown by 
the two copies of $\textbf{x}$ in Equation \eqref{eq:qhat}. If both copies of $\textbf{x}$ come from the same baseline and time, it 
can result in power spectrum measurements that are higher than those predicted by the thermal noise of the instrument. One 
way to avoid this type of noise bias is to avoid cross-multiplying data from the same baselines or days. This ensures that the 
two quantities that go into a measurement have separate noises that don't correlate with each other. We also note that if the noise level is known, this type of bias can be subtracted off, though this procedure is argued to be dangerous by \citet{dillon_et_al2014}.

@@

\begin{itemize}
\item \textbf{Null Test}: A null test is a type of jackknife test that removes the astronomical signal from data in order to 
investigate underlying systematics (e.g., see \citet{keating_et_al2016} for examples from intensity mapping that are closely related to our current application). For example, one can 
divide data into two subsets by separating odd and even Julian dates, or the first half of the observing season from the second. 
Subtracting the two removes signal that is common to both subsets, including foregrounds and EoR. The resulting power 
spectrum should be consistent with thermal noise estimates; if it is not, it suggests the presence of a systematic that differs 
from one of the data subsets to the other (i.e. doesn't get subtracted perfectly). 
\item \textbf{Traditional Jackknife}: In a broader sense, it is important to perform many jackknife tests in order to instill 
confidence in a final result. A stable result must be steadfast throughout all jackknives no matter how the data is sliced. 
Jackknives can be taken along several different axes --- for example, one could start with a full dataset, and compute a new 
power spectrum every time as a day of data is removed, or a baseline is removed. This type of jackknife would reveal bias 
present only at certain LSTs (such as a foreground source), for example, or misbehaving baselines.
\end{itemize}

>>>>

Bootstrapping uses sampling with replacement to estimate a posterior distribution. For example, measurements (like power 
spectra) can be made from different samples of data. Each of these measurements is a different realization drawn from some underlying distribution, and realizations are correlated with each other to a degree set by the fraction of sampled points that are held in common 
between them. Through the process of re-sampling and averaging along different axes, we can estimate error bars for 
our results which represent the underlying distribution of values that are allowed by our measurements (\citealt{efron_tibshirani1994}; \citealt{andrae2010}).

Suppose we have $N$ different measurements targeting the same quantity (for example, $N$ power spectrum measurements made along different axes, such as baselines or times). 
Bootstrapping means that we form $N_{boot}$ (often a large number) bootstraps, where each bootstrap is a random selection 
of the $N$ measurements. Bootstraps each have dimensions of $N$, and the values populated into each bootstrap are drawn 
from the original set of measurements with replacement (i.e. every $n^{th}$ slot in $N$ is filled randomly for each bootstrap). Next we take 
the mean of each bootstrap to collapse it from an array of length $N$ to a single number (we are interested in the mean statistic 
here, but any function of interest can be applied to each bootstrap as long as it's the same function for each one). The error (on 
the mean) is then computed as the standard deviation across all bootstraps.

@@

This example highlights the importance of understanding how analysis techniques (e.g. fringe-rate filtering) can affect a 
common statistical procedure like bootstrapping. Bootstrapping as a means of estimating power spectrum errors from real 
fringe-rate filtered data requires knowledge of the number of independent samples, which is not always a trivial task. For 
example, computing the effective number of independent samples of fringe-rate filtered data is not as simple as counting the 
number of averages performed. Down-sampling a time-averaged signal is straightforward using a boxcar average, but non-trivial with a more complicated convolution function that has long tails. Hence, we do not recommend bootstrapping unless the 
number of independent samples along the axis that is being re-sampled is well-determined. In Section \ref{sec:Boot}, we explain how we under-estimated errors in the \citetalias{ali_et_al2015} analysis of PAPER and how our bootstrapping procedure has now changed in light of this issue.

@@

@@

In Section \ref{sec:SiglossOverview}, we discussed signal loss as a form of multiplicative bias to estimates of the signal. Foregrounds are another type of bias, but an additive one instead of multiplicative. Foreground bias is perhaps one of the main factors limiting 21\,cm results, as foreground signals lie $\sim$$4$-$5$ orders of 
magnitude above the cosmological signal. Though there are many techniques proposed for removing foregrounds (see e.g. \citealt{vedantham_et_al2012}; \citealt{parsons_et_al2012a}; \citealt{parsons_et_al2012b}; \citealt{dillon_et_al2013a}; \citealt{wang_et_al2013}; \citealt{parsons_et_al2014}; \citealt{liu_et_al2014a}; \citealt{liu_et_al2014b}; \citealt{dillon_et_al2015}; \citealt{pober_et_al2016a}; \citealt{trott_et_al2016}), most 
experiments currently remain limited by residuals rather than noise, especially at low $k$.

@@

Although smooth spectrum foregrounds preferentially show up at low delay, or low $k$-modes, their isolation within the wedge is not perfect. In deep measurements, power spectrum measurements at $k_{\parallel}$ values beyond 
the delay associated with the length of a baseline are often still contaminated at a low level. This leakage, particularly at low $k$'s, can be attributed to 
convolution kernels associated with Fourier-transforming visibilities into delay-space. In other words, smooth-spectrum 
foregrounds appear as $\delta$-functions in delay-space, convolved by the Fourier transform of the source spectrum, the signal chain, and the 
antenna response, all of which could smear out the foregrounds and cause leakage outside the wedge \citep[e.g.][]{ewall-wice_et_al2017}.

@@

Determining the source of positive non-EoR detections at higher $k$'s is more difficult. In previous power spectrum results, these detections have been explained as instrumental systematics, particularly time-variable cross talk, RFI, cable reflections, and calibration errors (\citetalias{ali_et_al2015}; \citealt{parsons_et_al2014}; \citealt{dillon_et_al2014}; \citealt{beardsley_et_al2016}; \citealt{patil_et_al2017}). In the next section, we will present some tests that can help distinguish these excesses from that of EoR.

In addition to foreground bias, noise can also be responsible for positive power spectrum detections if thermal noise is 
multiplied by itself. Every 21\,cm visibility measurement contains thermal noise that is comprised of receiver and sky noise. 
We expect this noise to be independent between antennas and thus we can beat it down (increase sensitivity) by integrating 
longer, using more baselines, etc. However, the squaring of noise can occur when cross-multiplying visibilities, which is shown by 
the two copies of $\textbf{x}$ in Equation \eqref{eq:qhat}. If both copies of $\textbf{x}$ come from the same baseline and time, it 
can result in power spectrum measurements that are higher than those predicted by the thermal noise of the instrument. One 
way to avoid this type of noise bias is to avoid cross-multiplying data from the same baselines or days. This ensures that the 
two quantities that go into a measurement have separate noises that don't correlate with each other. We also note that if the noise level is known, this type of bias can be subtracted off, though this procedure is argued to be dangerous by \citet{dillon_et_al2014}.

@@

\begin{itemize}
\item \textbf{Null Test}: A null test is a type of jackknife test that removes the astronomical signal from data in order to 
investigate underlying systematics (e.g., see \citet{keating_et_al2016} for examples from intensity mapping that are closely related to our current application). For example, one can 
divide data into two subsets by separating odd and even Julian dates, or the first half of the observing season from the second. 
Subtracting the two removes signal that is common to both subsets, including foregrounds and the EoR signal. The resulting power 
spectrum should be consistent with thermal noise estimates; if it is not, it suggests the presence of a systematic that differs 
from one of the data subsets to the other (i.e. doesn't get subtracted perfectly). 
\item \textbf{Traditional Jackknife}: In a broader sense, it is important to perform many jackknife tests in order to instill 
confidence in a final result. A stable result must be steadfast throughout all jackknives no matter how the data is sliced. 
Jackknives can be taken along several different axes --- for example, one could start with a full dataset, and compute a new 
power spectrum every time as a day of data is removed, or a baseline is removed. This type of jackknife would reveal bias 
present only at certain LSTs (such as a foreground source), for example, or misbehaving baselines.
\end{itemize}
